Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-07-08 analysis/paper/article.qmd
got the title page sorted

---
title: "Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology"
author:
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - uw
  - Anne Marie Poole:
      institute: uw
      orcid: xxx
  - Ailin Zhang:
      institute: uw
      orcid: xxx
  - Setareh Shafizadeh:
      orcid: 0000-0003-2231-4310
      institute: uw
  - Jess Beck:
      orcid: 0000-0002-7387-2307
      institute:
        - ucd
institute:
  - uw: Department of Anthropology, University of Washington, Seattle, USA
  - ucd: School of Archaeology, University College Dublin, Ireland

title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 1000
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/society-for-american-archaeology.csl" # Insert path for the bib-style
abstract: |
  Many archaeology graduate students pursue advanced degrees in the hope of undertaking an academic career. Job listing websites often serve as the first port-of-call for students  seeking academic positions. We examined tenure-track job advertisements over the past decade to gain insights into the academic job market for archaeologists. Using data from the community-edited Academic Jobs Wiki for Archaeology, we examined changes in the academic job market over time. We investigated the editing dynamics of the Wiki to understand its users and their biases. We then analyzed the text of 431 job ads posted from 2013--2023. Our analysis addresses the question of how  archaeological topics, methods, and geographic regions specified in archaeological job ads have shifted over time. We also explored whether the labor burden for applicants has changed over time: do institutions request more information and documents from applicants at the initial stages of application, compared to a decade ago? Finally, we assessed the influence of socio-political factors on the changing focus of research topics in the field. We conclude with implications for archaeology students, graduates, and advisors seeking to understand the dynamics of the academic job market and the requirements of employers.
keywords: |
  archaeology; hiring; academia; job ads 
---

Keywords: `r rmarkdown::metadata$keywords`

@@

@@

```{r}
#| label: fig-panel-wiki-self-report-numbers
#| fig-cap: "Number of users self-reporting in each category of 'current users' per year"

@@

```{r}
#| label: fig-panel-wiki-characteristics
#| fig-cap: "A: Total number of edits to each Wiki page (additions and deletions) per year. B: Number of unique editors, as represented by distinct usernames or IP addresses (recorded for editors without usernames), active per year. C: Active periods of individual editors, the black lines connect the date of the first and last edit of all editors that were active for three months or more. D: Typical size of each edit per year, either addition or removal of text, one word is roughly 4-6 bytes. E: Distribution of the numbers of edits per job ad for each year. F: Histogram showing that the majority of editors are only active on the site for less than one year. G: Breakdown of self-reported Wiki user categories for each year."
#| fig-height: 4

@@

```{r}
#| label: fig-show-basic-plots
#| fig-cap: "A: total number of job ads from US institutions posted to the Academic Jobs Wiki for Archaeology in each year, with coloured sections showing the proportion of jobs by title and rank. B: Ratio of tenure-track to non-tenure-track positions over time. The red line indicates a 1:1 ratio of tenure-track to non-tenure-track positions, bars taller than that line indicate more tenure-track than non-tenure-track positions in that year"

@@

```{r}
#| label: fig-show-map-of-hiring-institution-and-cc
#| fig-cap: "A: Frequency of hiring institution by Carnegie classification. B: Inset shows map of the United States showing the count of tenure-track job ads posted by all institutions in each state during 2013 -- 2023"

@@

```{r}
#| label: fig-show-geo-trends
#| fig-cap: "A: Frequency of locations mentioned in the text of the job ads. B: Popularity of locations in job ads over time for locations that appear in 20 or more ads. Individual data points are shown, overlain by a locally weighted regression line for each location to indicate temporal trends."

@@

```{r}
#| label: fig-show-metho-trends
#| fig-cap: "A: Frequency of methods mentioned in the text of the job ads. B: Popularity of methods in job ads over time for methods that appear in 10 or more ads. Individual data points are shown, overlain by a locally weighted regression line for each location to indicate temporal trends."

@@

```{r}
#| label: fig-show-topi-trends
#| fig-cap: "A: Frequency of topics mentioned in the text of the job ads. B: Popularity of topics in job ads over time for topics that appear in 20 or more ads. Individual data points are shown, overlain by a locally weighted regression line for each location to indicate temporal trends."

@@

```{r}
#| label: fig-show-cooc
#| fig-cap: "Heatmap of topic co-occurrence in job ads."

@@

```{r}
#| label: fig-requirements-over-time
#| fig-cap: "Changing requirements in job ads over time. 'Count' refers to the number of each item required."

>>>>

---
# title: "Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology"
# author:
#   - Ben Marwick:
#       correspondence: "yes"
#       email: bmarwick@uw.edu
#       orcid: 0000-0001-7879-4531
#       institute:
#         - uw
#   - Anne Marie Poole:
#       institute: uw
#       orcid: xxx
#   - Ailin Zhang:
#       institute: uw
#       orcid: xxx
#   - Setareh Shafizadeh:
#       orcid: 0000-0003-2231-4310
#       institute: uw
#   - Jess Beck:
#       orcid: 0000-0002-7387-2307
#       institute:
#         - ucd
# institute:
#   - uw: Department of Anthropology, University of Washington, Seattle, USA
#   - ucd: School of Archaeology, University College Dublin, Ireland
#   
# title-block-published: "Last updated"  
# date: now
# date-format: long
# ---  Output Options ---
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 1000
filters:
 # - ../templates/scholarly-metadata.lua
 # - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/society-for-american-archaeology.csl" # Insert path for the bib-style
reference-section-title: "References Cited" # Required title for the reference list
---

@@

<!-- 
MANUAL SAA-COMPLIANT TITLE PAGE FOR WORD OUTPUT

For Word output, we create the title page manually in the document body.
The headings below will be styled using the "Heading 1" and "Normal"
styles from my reference docx in ../templates/template.docx
-->
\
\

::: {custom-style="Title Page Centered"}

Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

:::

<!-- Use backslashes for significant vertical spacing -->
\
\
<!-- This div applies our custom centered style to the top content -->

::: {custom-style="Title Page Centered not bold"}

<!-- This text will use the "Title Page Centered" style (Centered, Not Bold) -->
Ben Marwick, Anne Marie Poole, Ailin Zhang, Setareh Shafizadeh, and Jess Beck
\
DO NOT CITE IN ANY CONTEXT WITHOUT PERMISSION OF THE AUTHOR(S)
\

:::

\
\

::: {custom-style="Authors on title page"}
 **Ben Marwick**,  **Anne Marie Poole**,  **Ailin Zhang**,  **Setareh Shafizadeh** Department of Anthropology, University of Washington, Seattle, WA, USA (bmarwick@uw.edu, corresponding author)

 **Jess Beck** School of Archaeology, University College Dublin, Ireland

:::

<!-- This command creates a page break after the title page. -->
\newpage

Many archaeology graduate students pursue advanced degrees in the hope of undertaking an academic career. Job listing websites often serve as the first port-of-call for students  seeking academic positions. We examined tenure-track job advertisements over the past decade to gain insights into the academic job market for archaeologists. Using data from the community-edited Academic Jobs Wiki for Archaeology, we examined changes in the academic job market over time. We investigated the editing dynamics of the Wiki to understand its users and their biases. We then analyzed the text of 431 job ads posted from 2013--2023. Our analysis addresses the question of how  archaeological topics, methods, and geographic regions specified in archaeological job ads have shifted over time. We also explored whether the labor burden for applicants has changed over time: do institutions request more information and documents from applicants at the initial stages of application, compared to a decade ago? Finally, we assessed the influence of socio-political factors on the changing focus of research topics in the field. We conclude with implications for archaeology students, graduates, and advisors seeking to understand the dynamics of the academic job market and the requirements of employers.

Keywords: archaeology; hiring; academia; job ads

Muchos estudiantes de posgrado en arqueología cursan estudios avanzados con la esperanza de emprender una carrera académica. Los portales de ofertas laborales suelen ser el primer punto de referencia para quienes buscan puestos en el ámbito académico. Examinamos anuncios de plazas de planta (tenure-track) publicados durante la última década con el objetivo de comprender mejor el mercado laboral académico en arqueología. Utilizando datos del Academic Jobs Wiki for Archaeology, un recurso colaborativo editado por la comunidad, analizamos los cambios en dicho mercado a lo largo del tiempo. Investigamos también las dinámicas de edición del Wiki para comprender mejor a sus usuarios y los sesgos que puedan introducir. Posteriormente, analizamos el texto de 431 anuncios de empleo publicados entre 2013 y 2023. Nuestra investigación aborda cómo han variado, a lo largo del tiempo, los temas, métodos y regiones geográficas mencionados en las convocatorias para puestos académicos en arqueología. Asimismo, exploramos si la carga de trabajo para quienes postulan ha cambiado: ¿las instituciones solicitan hoy más documentos e información en las etapas iniciales del proceso, en comparación con hace una década? Finalmente, evaluamos la influencia de factores sociopolíticos en la evolución de los temas de investigación solicitados. Concluimos con una reflexión sobre las implicaciones de nuestros hallazgos para estudiantes, egresados y asesores que buscan comprender las dinámicas del mercado laboral académico y los requisitos de las instituciones empleadoras.

Palabras clave: arqueología; contratación académica; academia; convocatorias laborales

@@

```{r}
#| label: fig-panel-wiki-self-report-numbers
#| fig-cap: "Number of users self-reporting in each category of 'current users' per year"
#| fig-width: 3

@@

```{r}
#| label: fig-panel-wiki-characteristics
#| fig-cap: "A: Total number of edits to each Wiki page (additions and deletions) per year. B: Number of unique editors, as represented by distinct usernames or IP addresses (recorded for editors without usernames), active per year. C: Active periods of individual editors, the black lines connect the date of the first and last edit of all editors that were active for three months or more. D: Typical size of each edit per year, either addition or removal of text, one word is roughly 4-6 bytes. E: Distribution of the numbers of edits per job ad for each year. F: Histogram showing that the majority of editors are only active on the site for less than one year. G: Breakdown of self-reported Wiki user categories for each year."
#| fig-width: 3
#| fig-height: 2

@@

```{r, fig.width = 1, fig.height = 1}
#| label: fig-show-basic-plots
#| fig-cap: "A: total number of job ads from US institutions posted to the Academic Jobs Wiki for Archaeology in each year, with coloured sections showing the proportion of jobs by title and rank. B: Ratio of tenure-track to non-tenure-track positions over time. The red line indicates a 1:1 ratio of tenure-track to non-tenure-track positions, bars taller than that line indicate more tenure-track than non-tenure-track positions in that year"
#| fig-width: 1
#| fig-height: 1

@@

```{r}
#| label: fig-show-map-of-hiring-institution-and-cc
#| fig-cap: "A: Frequency of hiring institution by Carnegie classification. B: Inset shows map of the United States showing the count of tenure-track job ads posted by all institutions in each state during 2013 -- 2023"
#| fig-width: 3

@@

```{r}
#| label: fig-show-geo-trends
#| fig-cap: "A: Frequency of locations mentioned in the text of the job ads. B: Popularity of locations in job ads over time for locations that appear in 20 or more ads. Individual data points are shown, overlain by a locally weighted regression line for each location to indicate temporal trends."
#| fig-width: 3

@@

```{r}
#| label: fig-show-metho-trends
#| fig-cap: "A: Frequency of methods mentioned in the text of the job ads. B: Popularity of methods in job ads over time for methods that appear in 10 or more ads. Individual data points are shown, overlain by a locally weighted regression line for each location to indicate temporal trends."
#| fig-width: 3

@@

```{r}
#| label: fig-show-topi-trends
#| fig-cap: "A: Frequency of topics mentioned in the text of the job ads. B: Popularity of topics in job ads over time for topics that appear in 20 or more ads. Individual data points are shown, overlain by a locally weighted regression line for each location to indicate temporal trends."
#| fig-width: 3

@@

```{r}
#| label: fig-show-cooc
#| fig-cap: "Heatmap of topic co-occurrence in job ads."
#| fig-width: 3

@@

```{r}
#| label: fig-requirements-over-time
#| fig-cap: "Changing requirements in job ads over time. 'Count' refers to the number of each item required."
#| fig-width: 3
